---
title: "chaosrcpp: chaos, precision and recurrence"
author: "Andrii Patrikei"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{chaosrcpp: chaos, precision and recurrence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", dpi = 96, fig.align = "center")
has_gg <- requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("patchwork", quietly = TRUE)
has_mpfr <- requireNamespace("Rmpfr", quietly = TRUE)
has_plotly <- requireNamespace("plotly", quietly = TRUE)
has_desolve <- requireNamespace("deSolve", quietly = TRUE)
library(chaosrcpp)
```

## What this package is about

Three things that tend to live in separate corners of a curriculum sit
together here: the numerics of integrating a chaotic ODE, the concept of
sensitive dependence on initial conditions, and the tools used to quantify
the resulting time series. The compiled kernels make all of it fast enough to
play with interactively.

## 1. The Lorenz system in three precisions

The integrator is one C++ template. `precision = "fp32"` instantiates it
with `float`, `"fp64"` with `double`, and `"dd"` with a double-double type
that carries about 106 bits. `"mpfr"` runs the same RK4 formulas in the
arbitrary-precision Rmpfr package as a gold-standard check.

```{r lorenz}
tr <- lorenz(n = 8000, dt = 0.01)
tr
plot(tr, col = hcl.colors(nrow(tr), "Viridis"))
```

A million RK4 steps take a few hundredths of a second in `double`. The
double-double level costs only about five to seven times as much and is tens
of thousands of times faster than an interpreted Rmpfr loop, which is what
makes it the default high-precision reference:

```{r timing}
system.time(lorenz(n = 1e6, thin = 100))
system.time(lorenz(n = 1e6, thin = 100, precision = "dd"))
```

### The butterfly effect of rounding error

Everyone knows that two nearby initial conditions separate exponentially on
a chaotic attractor. Rounding error is a perturbation of the same kind: every
arithmetic operation nudges the state by a relative amount of about
$\epsilon$, the unit roundoff of the number format. `precision_divergence()`
runs the *same* initial condition, algorithm and step size at several
precisions and measures the distance to a high-precision reference.

```{r divergence, fig.height = 6.5, eval = has_gg}
d <- precision_divergence(t_max = 60)
d
plot_divergence(d)
```

The dashed lines are the back-of-the-envelope prediction
$\epsilon \, e^{\lambda_1 t}$ with $\lambda_1 \approx 0.906$ the largest
Lyapunov exponent. Single precision ($\epsilon \approx 6 \cdot 10^{-8}$)
loses the trajectory around $t \approx 18$, double precision
($\epsilon \approx 10^{-16}$) around $t \approx 40$. Neither run is "wrong":
both are legitimate shadows of the true dynamics, but they are not the same
shadow, and any claim about the state of a chaotic system at a specific time
has to reckon with this.

If Rmpfr is installed the reference can be the arbitrary-precision run, and
then the double-double level joins the comparison. That run is an R-level
loop and takes about a minute for `t_max = 60`, so it is shown in the README
rather than evaluated here.

```{r mpfr-check, eval = has_mpfr}
# Short cross-check: dd and mpfr agree to double resolution over 200 steps
m <- lorenz(n = 200, precision = "mpfr", mpfr_bits = 160)
dd <- lorenz(n = 200, precision = "dd")
max(abs(as.matrix(dd[, 2:4]) - as.matrix(m[, 2:4])))
```

### Interactive 3D view

```{r plotly, eval = has_plotly}
plot_attractor_3d(fp32 = lorenz(n = 2500, precision = "fp32"),
                  fp64 = lorenz(n = 2500))
```

### Lyapunov spectrum

The full spectrum comes from integrating a set of tangent vectors along the
trajectory and re-orthonormalising them by Gram-Schmidt after every step.
The exponents must add up to the trace of the Jacobian,
$-(\sigma + 1 + \beta) = -13.667$, which is a satisfying self-check.

```{r lyapunov}
lyapunov_spectrum("lorenz", n = 200000)
lyapunov_spectrum("roessler", dt = 0.02, n = 300000)
```

## 2. Maps

### Logistic map: bifurcation diagram and Lyapunov exponent

```{r logistic, fig.height = 6.5, eval = has_gg}
r <- seq(2.8, 4, length.out = 1500)
bif <- logistic_bifurcation(r)
ly <- logistic_lyapunov(r)
plot_bifurcation(bif, ly)
```

The shaded regions of positive exponent coincide exactly with the chaotic
bands of the diagram, and every periodic window inside the chaotic region is
a dip below zero. At $r = 4$ the exponent is $\log 2$:

```{r log2}
logistic_lyapunov(4, n = 1e5)$lambda - log(2)
```

```{r cobweb, fig.height = 5, fig.width = 5, eval = has_gg}
plot_cobweb(logistic_cobweb(3.9, x0 = 0.1, n = 80))
```

### Henon, Ikeda and the standard map

```{r henon-ikeda, fig.height = 4, fig.width = 8}
layout(matrix(1:2, 1))
plot(henon(n = 60000))
plot(ikeda(n = 80000), col = "#7b1e3a")
```

```{r standard-map, fig.height = 6.5, fig.width = 6.5}
plot(standard_map(K = 0.9, n_orbits = 60, n = 600))
```

## 3. Recurrence plots and RQA

A recurrence plot marks the pairs of times $(i, j)$ at which the state
returns to within $\epsilon$ of itself. Its line structures carry a
surprising amount of information about the dynamics, and Recurrence
Quantification Analysis turns them into a handful of numbers: the recurrence
rate (RR), the fraction of points on diagonal lines (determinism, DET), the
average and longest diagonal line (L, Lmax), the entropy of the line length
distribution (ENTR), and the vertical-line analogues laminarity (LAM) and
trapping time (TT).

```{r rp, fig.height = 4.2, fig.width = 8.4}
tr <- lorenz(n = 3000, dt = 0.02)
layout(matrix(1:2, 1))
plot(tr)
plot_recurrence(tr, rr = 0.05)
```

```{r rqa}
rqa(tr, rr = 0.05)
```

The same measures separate a periodic signal from noise at a glance:

```{r rqa-compare}
set.seed(1)
periodic <- sin(seq(0, 30 * pi, length.out = 1500))
noise <- rnorm(1500)
rbind(periodic = rqa_features(periodic, m = 3, tau = 10),
      noise    = rqa_features(noise, m = 3, tau = 1))
```

`rqa_features()` returns one row per signal, which is the shape you want
when RQA measures are fed into a classifier or a clustering method as
features. The delay embedding for scalar series is done with `embed_delay()`
(or the `m` and `tau` arguments); multivariate input such as a
`chaos_trajectory` is used as-is.

Definitions follow Marwan, Romano, Thiel and Kurths (2007), *Recurrence plots
for the analysis of complex systems*, Physics Reports 438, 237-329. The
diagonal and vertical line histograms are returned too (`$diag_hist`,
`$vert_hist`) for anyone who wants to build other measures on top.

## 4. Benchmark

```{r bench, eval = has_desolve}
benchmark_lorenz(n = 100000)
```

All four implementations run the identical scheme, so they agree to rounding
over the comparison window. The speed difference is the whole reason the
compiled kernel exists: it turns "run the divergence experiment once" into
"sweep it over a grid of parameters".

## 5. Interactive explorer

```{r shiny, eval = FALSE}
run_chaos_explorer()
```

The Shiny app exposes the Lorenz parameters, the fp32-versus-fp64 comparison,
the logistic map with a movable `r` and the standard map with adjustable kick
strength.
